Something about Performance, 2D Component of NEMO. (TBC)

### 2015-02-12 nemolite2d_perf.tex
Text describing reasoning behind in-lining and SIMD directives

\subsection{Addition of `safe-loop' directives}

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the nemolite2d kernels contain conditional statements. These
statements are there to check whether {\it e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all arrays holding field data with extents greater
than strictly required such that any array accesses within the kernels
would remain within bounds.

@@

@@

@@

>>>>

\subsection{Addition of `safe-address' directives}

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the nemolite2d kernels contain conditional statements. These
statements are there to check whether {\it e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the safe\_address
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

@@

From our previous work~\cite{shallow_psykal} on a similar code we know
that kernel in-lining is critical to obtaining performance with both
the Gnu and Intel compilers. For the Gnu compiler, this is because it
cannot perform in-lining when routines are in separate source
files. In our previous work we obtained an order-of-magnitude speed-up
simply by moving subroutines into the module containing the middle
layer (from which the kernels are called). A further performance
improvement of roughly 30\% was obtained when the kernel code was
manually inserted at the site of the subroutine call.

Although the Intel compiler does do in-lining when routines are in
separate source files, we have found (both here and in our previous
work~\cite{shallow_psykal}) that the number of optimisations it
performs is reduced if it first has to in-line a routine. For the
Intel-compiled Shallow code, manually inserting kernel code at the
site of the subroutine call increased performance by about 25\%.

In fact, in-lining can have a significant effect on the Intel
compiler's ability to vectorise a loop. Taking the loop that calls the
kernel for the u-component of momentum as an example, before in-lining
the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(85,7) 
  inlined into nemolite2d.f90(86,11)
 remark #15335: loop was not vectorized: 
  vectorization possible but seems 
  inefficient. 
 --- begin vector loop cost summary ---
 scalar loop cost: 1307 
 vector loop cost: 2391.000 
 estimated potential speedup: 0.540 
 lightweight vector operations: 1133 
 medium-overhead vector operations: 1 
 vectorized math library calls: 1 
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
After we have manually in-lined the kernel body, the compiler reports:
\begin{verbatim}
LOOP BEGIN at time_step_mod.f90(97,7) 
  inlined into nemolite2d.f90(86,11)
 LOOP WAS VECTORIZED
 --- begin vector loop cost summary ---
 scalar loop cost: 1253 
 vector loop cost: 521.750 
 estimated potential speedup: 2.350 
 lightweight vector operations: 697 
 medium-overhead vector operations: 1 
 vectorized math library calls: 1 
 --- end vector loop cost summary ---
LOOP END
\end{verbatim}
Looking at the `estimated potential speedup' in the compiler outputs
above, it is clear that the way in which the compiler vectorises the
two versions must be very different. This conclusion is borne out by
the fact that if one persuades the compiler to vectorise the first
version (through the use of a directive) then the performance of the
resulting binary is worse than that where the loop is left
un-vectorised.

The profiling data in Table~\ref{TABLE_profile} shows that it is the
Momentum section that accounts for the bulk of the model run-time.  We
therefore chose to in-line the two kernels (one for each of the $x$
and $y$ components of momentum) called within this section. Given our
findings with the shallow code, we have chosen to do full, manual
inlining of the kernel code.

@@

It turns out that the Cray-compiled binaries of both the original and
\psykal versions of nemolite2d perform considerably less well than
their Intel-compiled counterparts. Comparison of the diagnostic output
from each of the compilers revealed that while the Intel compiler was
happy to vectorise the Momentum loops, the Cray compiler was choosing
not to:
\begin{verbatim}
   99.          do ji = 2, M-1, 1
 A loop starting at line 99 was blocked 
 with block size 256.

 A loop starting at line 99 was not
 vectorized because it contains 
 conditional code which is more
 efficient if executed in scalar mode.
\end{verbatim}
Inserting the Cray `vector always' directive persuaded the compiler
to vectorise the loop:
\begin{verbatim}
  99.    !dir$ vector always
  100.          do ji = 2, M-1, 1
 A loop starting at line 100 was blocked
 with block size 256.

 A loop starting at line 100 requires an 
 estimated 17 vector registers at 
 line 151; 1 of these have been 
 preemptively forced to memory.

 A loop starting at line 100 was vectorized.
\end{verbatim}
which gave a significant performance improvement. This behaviour is in
contrast to that obtained with the Intel compiler: its predictions
about whether vectorising a loop would be beneficial were generally
found to be reliable.

@@

Further comparison of the diagnostic output from the Cray and Intel
compilers revealed that the Cray compiler was vectorising the
Continuity section while the Intel compiler reported that it was
unable to do so due to dependencies. After some experimentation we
found that this was due to the compiler's limited analysis of the way
components of Fortran derived types were being used.
